Working from home and digital divides: resilience during the pandemic

### 2021-03-25 paper/v2_taylor_francis/broadband.speed.covid.tex
new appendix with 2019 clusters, text for discussion, new images of higher res, ref and footnote for openair added

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds during the spring 2020 lockdown, may
reinforce or redress the spatial and social dimensions of digital
divisions. We employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. As the quality of
internet infrastructure and services, as well as the concentration of
different occupations are spatially dependent and clustered in space,
our approach offers a framework for understanding the impact of and
interactions between digital divisions geographically and
socioeconomically. Our analysis enables us to better understand how the
spatial and social distribution of education, occupation and online
accessibility intersect to enable or hinder the practice of
telecommuting at a time of extreme demand. We also consider what lessons
can be learned from this time for a future where telecommuting is likely
to remain a more common means of accessing work, at least in comparison
to the pre-Covid era.

@@

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
In some aspects, these returns may have increased during the Covid-19
crisis, when those with the capability to telecommute also have the
ability to maintain their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the
first level digital divide, which is associated with access to and
quality of internet connectivity. \citet{SALEMINK2017360} provides a
systematic review of the pre-pandemic, first level digital divide in
infrastructure quality between urban and rural areas in various advanced
economies. Rural areas, predictably, fare worse. This finding broadly
applies in the UK, where studies that also analysed broadband speed
checks concluded that average speeds are lower in rural areas, something
that has not been improved by policy measures to increase competition
\citep{riddlesden2014broadband, nardotto2015unbundling}. Moreover,
\citet{riddlesden2014broadband} found that levels of deprivation did not
correlate with first level digital divides.

However, as \citet{blank2018local} highlight, variation in individual
internet uptake and use is a product of more complex spatial and
demographic characteristics than simple rurality or urbanisation. Dense
urban areas were shown to suffer more from slowdown during peak hours,
although these services were more likely to be improved by increased
competition between providers, such as between new entrants and Virgin
Media cable connections
\citep{riddlesden2014broadband, nardotto2015unbundling}. The latter were
historically available to only \(45\)\% of premises in the UK
\citep{ofcom2016}, where the more lucrative and competitive market
originally attracted the cable TV provider. Whether the variation in
infrastructure quality and reliability affects the spatial footprint of
telecommuting has not previously been investigated, in part because
telecommuting has not previously been a cause of concentrated demand and
pressure on internet services. Other intersections between digital
divides have been subject to study. There are indications that those who
purchase high speed connections consume more data of all sorts and use
their connections for a greater variety of purposes
\citep{hauge2011consumer}. There is also a correlation between access to
internet services and a reduction in household transport spend
\citep{bris2017ict}. These studies suggest that better internet services
enable households to make savings and efficiencies, an example of the
first level digital divide reinforcing the third level.

Currently there is a hotly debated discussion in the literature
regarding these exact changes in transportation and the structure of
cities. We now have enough hard data to observe the drastic change in
the space-time geography of cities around the world
\citep[e.g.~][]{google2020, shibayama2021impact}. Although there is a
broad agreement that these changes during the pandemic have played a
pivotal role in stopping the spread of the virus
\citep{jia2020population, yang2020taking, mu2020interplay}, we still
have no evidence to what extend these increased levels of working from
home, and the consequent decrease in commuting flows and the structure
of cities, will remain post-pandemic.

Multi-layered digital divides may also intersect with material divides
and the economic geography of the UK. The regional economic resilience
literature underlines the differentiated capacity of cities and regions
to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}. As different places have
different industrial and occupational profiles, these affect their
aggregated potential capacity for telecommuting. Such profiles are
associated with longstanding inequalities in the UK and their spatial
representation as a North-South divide \citep{martin_north_south}.
Various studies have illustrated severe inequalities between the north
and the south regions of England in terms of skills and human capital,
unemployment, productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers from some of the highest
levels of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. All three levels of digital
divides are associated with or shaped by the geography of the UK. Yet
this is the first time that the intersection of digital and material
divides is relevant to understanding the economic resilience of places
and large swathes of the population, as digital technologies became an
essential tool of productivity during the Covid-19 pandemic.

The extreme demand during the pandemic thus provides a new opportunity
to understand how internet infrastructure quality, and reliability
affects telecommuting, particularly in light of the high volumes of
bandwidth-intensive video conferencing required in order to avoid the
face-to-face contact that could increase the spread of infection. We
seek to answer how internet service resilience could contributes to or
reduce economic resilience when the latter is dependent upon the
capability to work from home. We also aim to improve our understanding
of the impact of first level digital divisions on telecommuting, and
whether this results in more fundamental third level digital divisions
than has previously been perceived.

@@

The frequency of speed tests was important in identifying the temporal
profile which would give us most insight into experienced internet
service and resilience over units of time. Whilst there is an overall
trend of increased testing from March to April and then a slight
reduction from April to May, this trend masks substantial variation by
not only the day of the week, but also time of day, as can be seen in
Figure \ref{test2020}. Thus, a daily aggregation of upload speeds would
mask the variation in experienced service over the course of each
weekday. Furthermore, the importance of this variation is highlighted by
a comparison with the same period in \(2019\), as in Figure
\ref{test2019}, when the volume of testing and thus of experienced
internet service quality peaked in the evening, presumably in response
to demand for leisure activities and download speeds. In contrast, the
majority of the increase in testing in \(2020\) is during the working
day, creating a new morning peak in Figure \ref{test2020}. Therefore, we
include a measure of hourly variation in our temporal profiles to
reflect the change in users' perception of the workday reliability of
internet services.

@@

@@

First, we control for the number of speed tests run per cluster
inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
Virgin Media internet connections. Regarding the former, we expect
people in LADs with more unreliable connections to test their internet
speeds more often, and the results show that those in cluster \(6\) are
by far the most concerned about reliability at that time of day.
Meanwhile, those in clusters \(2\), \(3\), and \(9\) benefit from a
higher proportion of Virgin connections, which is an indication that
people in these clusters are more likely to live in urban areas, with
more profitable broadband markets, and a better choice of broadband
services. We also employ distance from London and from the nearest
metropolitan area (including London) as two variables depicting
peripherality and urban structure. However, whilst significant, the
pattern of coefficients of these variables is inconclusive in confirmed
first level digital divides. Even though London was one of the ten
largest metropolitan areas in England, which, along with Glasgow and
Cardiff, were identified to calculate the variable estimating the impact
of distance from the centre of a metropolitan area, the coefficients for
the two variables are mostly small and the signs for some are opposite.

@@

In summary, the regression results indicate the complexity of measuring
second and third level digital divides within spatial aggregations where
the geography of first level digital divides has been captured using
time-series clustering of experienced broadband upload speeds as a
product of reliability not just availability. Whilst the differences in
mean speeds between clusters were not large, the temporal clustering of
internet resilience showed much greater variation, and was not spatially
dependent upon distance to large urban areas, or relative location.
Internet resilience supports a wide range of small and large urban
economies, but also fails, or at least frustrates a wide range of other
urban economies, as demonstrated by the number of tests per capita run
in cluster \(6\). Yet our analysis of the likelihood of furlough within
the population, especially in the smaller suburban clusters \(7\),
\(3\), \(5\), and also \(8\) shows that digital divides were as likely
to diverge as intersect during the pandemic, and did not necessarily
overlap with prior economic divides. With a much greater share of the
population likely to continue to work from home in the future, and with
changing attitudes towards residential locations, our analysis suggests
that first level digital divides should be seen as a function of
reliability as well as availability, of upload speeds as well as
download speeds, and of greater spatial variation than might have
previously been considered.

@@

This paper offers a new perspective on telecommuting from the viewpoint
of the complex web of digital divides. We employ novel data regarding
experienced upload speeds and time-series clustering methods, a family
of unsupervised machine learning techniques which are rarely utilised in
geographical research. Fast, reliable internet connections are necessary
for the population to be able to work from home. Although not every
place hosts individuals in occupations which allow for telecommuting nor
with the necessary skills to effectively use the internet to
telecommute, this paper raises the issue that places may depend upon
good internet reliability as well as connectivity to achieve economic
resilience in a period like the current pandemic when internet
resilience is so vital.

Our analysis demonstrated that the temporal profiles of seven of our
nine clusters had slower upload speeds in the morning than in the
evening. The opposite is likely to have been the norm prior to the
pandemic, as level of demand and bandwidth management is the most common
cause of temporal variation in experienced speeds, and why evening
download speeds, rather than daytime upload speeds, have been used to
benchmark the performance of internet services. Thus, the new patterns
can be taken as evidence of widespread telecommuting and other daytime
internet use which changed the temporal profile of internet activity
throughout the UK, not just in areas with more digital industry or
better skills. Furthermore, upload speeds have not previously been seen
as integral to universal service, considering there has never before
been such extreme demand for telecommuting and operations such as video
calls. Yet those in the largest cluster, \(6\), clearly experienced
lower upload speeds. Their speeds were also less reliable, perhaps
because they had higher levels of demand than the second largest
cluster, \(9\), where many more employees were furloughed.

Almost all of the largest urban areas in the UK, as well as many smaller
cities, were split between the largest two clusters -- \(6\) and \(9\)
-- and both contain LADs that are centres of the knowledge economy, from
Cambridge in the former to Oxford in the latter, or have high
concentrations of digital businesses, like Reading in the former and
Milton Keynes in the latter \citep{technation2017}. Yet there are both
intersections and divergences of digital and economic divides. LADs in
cluster \(9\) were able to benefit both from reliable internet
connections and populations which were familiar with working from home
and could capitalise on their digital infrastructure, although overall,
they appeared to have a lower rate of return in the pandemic, with far
greater numbers furloughed. Is this an indication of both historic
economic division, as well as the third level digital divide in cities
in the North and Midlands unable to capitalise on their digital
infrastructure? Meanwhile, did the greater presence in London and the
South of LADs in cluster \(6\), as well as the inclusion of many rural
areas less impacted by the pandemic help those in this cluster be more
resilient and gain greater returns, despite using less reliable internet
services? The nuanced picture we gained through our analysis of the UK
case study suggests that in cluster \(1\) as well as \(6\), being on the
right side of the second level digital divide had a greater impact on
economic resilience and therefore the third level digital divide, than
having quality internet connectivity.

Digital infrastructure which considers upload speeds and working day
reliability as well as availability are likely to be particularly
important in a future where telecommuting might be a more common means
of accessing work and broadband services must be fit for purpose,
although the long-term effects of such drastic changes in telecommuting
and attitudes towards working from home are difficult to predict.
Nevertheless, they span various aspects of economy and society: from
changes to transportation planning due to altered commuting patterns to
changes in land use and urban planning to accommodate people who work
from home \citep{BUDNITZ2020102713, ELLDER2020102777}, and from
productivity and innovation changes to changes in agglomeration
externalities and the attraction of large cities \citep{econobs}.
Further research may be able to measure the economic resilience of the
different clusters of places discussed in this paper once this pandemic
is firmly past. However, our analysis demonstrates that the economic
resilience made possible by working from home cannot be understood
without considering the underpinning digital divides and cannot be
achieved without planning for how the levels of digital, social and
economic divides might intersect.

Early research on Covid-19 and cities speculated the changes that
potential extensive post-Covid working from home patterns might generate
for spatial structure: from more walkable cities to more localised
production and consumption patterns and from more extensive urban sprawl
to decline in public transportation and increased private car usage
\citep{batty2020editorial}. Despite the essential role it played during
the pandemic, less effort has been spent in understanding the current
and future role of digital infrastructure. Our results indicate that
probably for the first time the future of cities and spatial structure
are so much intertwined with digital infrastructure. If the post-covid
world is a world with extensive working from home, then we need to
build, among other things, resilient digital infrastructure capable of
bridging the first layer of the digital divide. Contrary to previous
broadband business and deployment plans, emphasis should be placed not
only on download, but also on upload speeds {[}\citet{brake2020lessons}.
In essence, Covid-19 accentuated the old, but more valid than ever
argument: digital infrastructure, just like any other network
infrastructure, only becomes visible when it stops working
\citep{star1999, tranos2013geography}.

@@

>>>>

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds during the spring 2020 lockdown, may
reinforce or redress the spatial and social dimensions of digital
divisions. We employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. As the quality of
internet infrastructure and services, as well as the concentration of
different occupations are spatially dependent and clustered in space,
our approach offers a framework for understanding the impact of and
interactions between digital divisions geographically and
socioeconomically. Our analysis enables us to better understand how the
spatial and social distribution of occupation and online accessibility
intersect to enable or hinder the practice of telecommuting at a time of
extreme demand. We also consider what lessons can be learned from this
time for a future where telecommuting is likely to remain a more common
means of accessing work, at least in comparison to the pre-Covid era.

@@

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
In some aspects, these returns may have increased during the Covid-19
crisis, when those with the capability to telecommute also have the
ability to maintain their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the
first level digital divide, which is associated with access to and
quality of internet connectivity. \citet{SALEMINK2017360} provides a
systematic review of the pre-pandemic, first level digital divide in
infrastructure quality between urban and rural areas in various advanced
economies. Rural areas, predictably, fare worse. This finding broadly
applies in the UK, where studies that also analysed broadband speed
checks concluded that average speeds are lower in rural areas, something
that has not been improved by policy measures to increase competition
\citep{riddlesden2014broadband, nardotto2015unbundling}. In contrast,
\citet{riddlesden2014broadband} found that levels of deprivation did not
correlate with first level digital divides.

However, as \citet{blank2018local} highlight, variation in individual
internet availability and uptake is a product of more complex spatial
and demographic characteristics than simple rurality or urbanisation.
Dense urban areas were shown to suffer more from slowdown during peak
hours, although these services were more likely to be improved by
increased competition between providers, such as between new entrants
and Virgin Media cable connections
\citep{riddlesden2014broadband, nardotto2015unbundling}. The latter were
historically available to only \(45\)\% of premises in the UK
\citep{ofcom2016}, where the more lucrative and competitive market
originally attracted the cable TV provider. Whether the variation in
infrastructure quality and reliability affects the spatial footprint of
telecommuting has not previously been investigated, in part because
telecommuting has not previously been a cause of concentrated demand and
pressure on internet services.

Other intersections between digital divides have been subject to study.
There are indications that those who purchase high speed connections
consume more data of all sorts and use their connections for a greater
variety of purposes \citep{hauge2011consumer}. There is also a
correlation between access to internet services and a reduction in
household transport spend \citep{bris2017ict}. These studies suggest
that better internet services enable households to make savings and
efficiencies, an example of the first level digital divide reinforcing
the third level. Such digital intersections and efficiencies have also
changed transportation, connectivity and the structure of cities during
the pandemic. We now have enough data to confirm the drastic alteration
observed in the space-time geography of cities around the world in the
last year \citep[e.g.~][]{google2020, shibayama2021impact}. Yet,
although there is a broad agreement that these changes during the
pandemic have played a pivotal role in stopping the spread of the virus
\citep{jia2020population, yang2020taking, mu2020interplay}, the extent
to which the increased levels of working from home, the consequent
decrease in commuting flows, and the altered structure of cities will
remain post-pandemic is the subject of considerable debate in the
literature.

Meanwhile, multi-layered digital divides may also intersect with
material divides and the economic geography of the UK. The regional
economic resilience literature underlines the differentiated capacity of
cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}. As different places have
different industrial and occupational profiles, these affect their
aggregated potential capacity for telecommuting. Such profiles are
associated with longstanding inequalities in the UK and their spatial
representation as a North-South divide \citep{martin_north_south}.
Various studies have illustrated severe inequalities between the north
and the south regions of England in terms of skills and human capital,
unemployment, productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers from some of the highest
levels of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. All three levels of digital
divides are associated with or shaped by the geography of the UK. Yet
this is the first time that the intersection of digital and material
divides is relevant to understanding the economic resilience of places
and large swathes of the population, as digital technologies became an
essential tool of productivity during the Covid-19 pandemic.

The extreme demand during the pandemic thus provides a new opportunity
to understand how internet infrastructure quality, and reliability
affects telecommuting, particularly in light of the high volumes of
bandwidth-intensive video conferencing required in order to avoid the
face-to-face contact that could increase the spread of infection. We
seek to answer how internet service resilience could contribute to or
reduce economic resilience when the latter is dependent upon the
capability to work from home. We also aim to improve our understanding
of the impact of first level digital divisions on telecommuting, and
whether this results in more fundamental third level digital divisions
than has previously been perceived.

@@

The frequency of speed tests was important in identifying the temporal
profile which would give us most insight into experienced internet
service and resilience over units of time. Whilst there is an overall
trend of increased testing from March to April and then a slight
reduction from April to May, this trend masks substantial variation by
not only the day of the week, but also time of day, as can be seen in
Figure \ref{test2020} \footnote{Figures \ref{test2020} and
  \ref{test2019} were created with the \texttt{R} package
  \texttt{openair} \citep{openair}.}. Thus, a daily aggregation of
upload speeds would mask the variation in experienced service over the
course of each weekday. Furthermore, the importance of this variation is
highlighted by a comparison with the same period in \(2019\), as in
Figure \ref{test2019}, when the volume of testing and thus of
experienced internet service quality peaked in the evening, presumably
in response to demand for leisure activities and download speeds. In
contrast, the majority of the increase in testing in \(2020\) is during
the working day, creating a new morning peak in Figure \ref{test2020}.
Therefore, we include a measure of hourly variation in our temporal
profiles to reflect the change in users' perception of the workday
reliability of internet services.

@@

It is worth noting here that clusters of LAD based on the upload speeds
are very different to the ones occur when we apply the same methods to
upload speed data for the same time period in 2019 -- see Appendix
\protect\hyperlink{appendix2}{2}. This drastic difference between the
LAD upload speed profiles before and during the pandemic is an
indication of the changes in internet usage that took place during the
pandemic.

@@

First, we control for the number of speed tests run per cluster
inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
Virgin Media internet connections \footnote{See Appendix
  \protect\hyperlink{appendix3}{3} for the descriptive statistics.}.
Regarding the former, we expect people in LADs with more unreliable
connections to test their internet speeds more often, and the results
show that those in cluster \(6\) are by far the most concerned about
reliability at that time of day. Meanwhile, those in clusters \(2\),
\(3\), and \(9\) benefit from a higher proportion of Virgin connections,
which is an indication that people in these clusters are more likely to
live in urban areas, with more profitable broadband markets, and a
better choice of broadband services. We also employ distance from London
and from the nearest metropolitan area (including London) as two
variables depicting peripherality and urban structure. However, whilst
significant, the pattern of coefficients of these variables is
inconclusive in confirmed first level digital divides. Even though
London was one of the ten largest metropolitan areas in England, which,
along with Glasgow and Cardiff, were identified to calculate the
variable estimating the impact of distance from the centre of a
metropolitan area, the coefficients for the two variables are mostly
small and the signs for some are opposite.

@@

In summary, the regression results indicate the complexity of measuring
second and third level digital divides within spatial aggregations where
the geography of first level digital divides has been captured using
time-series clustering of experienced broadband upload speeds as a
product of reliability not just availability. Whilst the differences in
mean speeds between clusters were not large, the temporal clustering of
internet resilience showed much greater variation, and was not spatially
dependent upon distance to large urban areas, or relative location.
Internet resilience supports a wide range of small and large urban
economies, but also fails, or at least frustrates a wide range of other
urban economies, as demonstrated by the number of morning peak tests per
capita run in cluster \(6\). Yet our analysis of the likelihood of
furlough within the population, especially in the smaller suburban
clusters \(7\), \(3\), \(5\), and \(8\) shows that digital divides were
as likely to diverge as intersect during the pandemic, and did not
necessarily overlap with prior economic divides. With a much greater
share of the population likely to continue to work from home in the
future, and with changing attitudes towards residential locations, our
analysis suggests that first level digital divides should be seen as a
function of reliability as well as availability, of upload speeds as
well as download speeds, and of greater spatial variation than might
have previously been considered.

@@

Our analysis demonstrated that the temporal profiles of seven of our
nine clusters had slower upload speeds in the morning than in the
evening. The opposite is likely to have been the norm prior to the
pandemic, as level of demand and bandwidth management is the most common
cause of temporal variation in experienced speeds, and why evening
download speeds, rather than daytime upload speeds, have been used to
benchmark the performance of internet services. Thus, the new patterns
can be taken as evidence of widespread telecommuting and other daytime
internet use which changed the temporal profile of internet activity
throughout the UK, not just in areas with more digital industry or
better skills. Furthermore, upload speeds have not previously been seen
as integral to universal service, considering there has never before
been such extreme demand for telecommuting and operations such as video
calls. Yet those in the largest cluster, \(6\), clearly experienced
lower upload speeds. Their speeds were also less reliable, perhaps
because they had higher levels of demand than the second largest
cluster, \(9\), where many more employees were furloughed. Indeed, those
with the skills and jobs to work from home were often left with the
least robust services and greatest slowdown.

Therefore, home-based digital infrastructure which considers upload
speeds and working day reliability as well as availability are likely to
be particularly important in a future where telecommuting might be a
more common means of accessing work and broadband services must be fit
for purpose. Although the long-term effects of such drastic changes in
telecommuting and attitudes towards working from home are difficult to
predict, the reliability of home broadband services deserve more
consideration than in the past. This would represent a switch from
previous demand-side broadband policies in the UK which tended to be
aimed at supporting small and medium enterprises
\citep{HENDERSON2020102024}. SUch policies were based on previous
research regarding the productivity effects of broadband infrastructure
\citep{DESTEFANO2018110}. However, this stream of research tended not to
consider residential locations as places that host economic activity.
Such policies also made some assumptions about the quality of service in
residential markets that could be assured without government
intervention -- an assumption which our analysis suggest is not entirely
accurate.

Policy and development proposals should also consider how the potential
changes wrought by the pandemic span various aspects of economy and
society. Changes to transportation planning due to altered commuting
patterns also imply changes in land use and urban planning to
accommodate people who work from home
\citep{BUDNITZ2020102713, ELLDER2020102777}. Productivity and innovation
changes will reflect upon changes in agglomeration externalities and the
attraction of large cities \citep{econobs}. Further research may be able
to measure the economic resilience of the different clusters of places
discussed in this paper once this pandemic is firmly past. However, our
analysis demonstrates that the economic resilience made possible by
working from home cannot be understood without considering the
underpinning digital divides and cannot be achieved without planning for
how the levels of digital, social and economic divides might intersect.
For example, our results suggest that broadband policies cannot improve
the economic resilience of places where the industrial structure does
not align with occupations that incorporate the digital skills and
capabilities to work from home. Such places instead experienced higher
proportions of their labour force being placed on furlough.

Early research on Covid-19 and cities also speculated upon the changes
that potential extensive post-Covid working from home patterns might
generate for spatial structure: from more walkable cities and more
localised production and consumption patterns to more extensive urban
sprawl, the decline in public transportation and increased private car
usage \citep{batty2020editorial}. Despite the essential role it played
during the pandemic, less effort has been spent in understanding the
current and future role of digital infrastructure. Our results indicate
that probably for the first time the future of cities and spatial
structure are so intertwined with digital infrastructure. If the
post-covid world is a world with extensive working from home, then we
need to build, among other things, resilient digital infrastructure
capable of bridging the first layer of the digital divide. Contrary to
previous broadband business and deployment plans, emphasis should be
placed not only on download, but also on upload speeds
\citep{brake2020lessons}. In essence, Covid-19 accentuated the old
argument, now more valid than ever: digital infrastructure, just like
any other network infrastructure, only becomes visible when it stops
working \citep{star1999, tranos2013geography}.

On the other hand, the nuanced picture we gained through our analysis of
the UK case study suggests that being on the right side of the second
level digital divide had a greater impact on economic resilience and
therefore the third level digital divide, than having quality internet
connectivity. Our regressions results for cluster \(1\) and \(6\) are a
demonstration of this. Our analysis shows both the intersections and
divergences of digital and economic division. Almost all of the largest
urban areas in the UK, as well as many smaller cities, were split
between the largest two clusters -- \(6\) and \(9\) -- and both contain
LADs that are centres of the knowledge economy, from Cambridge in the
former to Oxford in the latter, or have high concentrations of digital
businesses, like Reading in the former and Milton Keynes in the latter
\citep{technation2017}. Yet whilst LADs in cluster \(9\) were able to
benefit both from reliable internet connections and populations familiar
with working from home to capitalise on their digital infrastructure,
they appeared to have a lower rate of return in the pandemic, with far
greater numbers furloughed. Is this an indication of historic economic
division, as well as the third level digital divide, in cities in the
North and Midlands unable to capitalise on their digital infrastructure?
Meanwhile, did the greater presence in London and the South of LADs in
cluster \(6\), as well as the inclusion of many rural areas less
impacted by the pandemic help those in this cluster be more resilient
and gain greater returns, despite using less reliable internet services?
Further research would be necessary to prise apart the detail,
particularly in the largest clusters where averages are less distinct.

In conclusion, this paper offers a new perspective on telecommuting from
the viewpoint of the complex web of digital divides. We employ novel
data regarding experienced upload speeds and time-series clustering
methods, a family of unsupervised machine learning techniques which are
rarely utilised in geographical research. Fast, reliable internet
connections are necessary for the population to be able to work from
home. Although not every place hosts individuals in occupations which
allow for telecommuting nor with the necessary skills to effectively use
the internet to telecommute, this paper raises the issue that places may
depend upon good internet reliability as well as connectivity to achieve
economic resilience in a period like the current pandemic when internet
resilience is so vital.

\pagebreak

@@

\begin{figure}
\includegraphics[width=1\linewidth]{figures/map.up.2019.clusters} \caption{\label{map.up.2019.clusters}Upload speed clusters for LADs in before the pandemic (2019)}\label{fig:unnamed-chunk-17}
\end{figure}

\pagebreak

\hypertarget{appendix3}{%
\section*{Appendix 3}\label{appendix3}}
\addcontentsline{toc}{section}{Appendix 3}
